Towards open collaboration in biomedical and health research: An open collaborative scoping review

### 2025-03-18 doc/paper.qmd
chore: :wrench: add Mario's ORCID

---
title: "Towards open collaboration in biomedical and health research: An open collaborative scoping review"
date: last-modified
author:
  - name: Mario Garcia
    affiliations:
      - ref: ku
  - name: Daniel B. Ibsen
    affiliations:
      - ref: sdca
      - ref: au
  - name: Luke W. Johnston
    affiliations:
      - ref: sdca
      - ref: au
affiliations:
  - id: sdca
    name: Steno Diabetes Center Aarhus
    country: Denmark
  - id: au
    name: Aarhus University
    country: Denmark
  - id: ku
    name: Copenhagen University
    country: Denmark
format:
  html: default
  academic-typst:
    section-numbering: "1.1.1"
    columns: 2
    papersize: a4
    margin:
      x: 2cm
      y: 2cm
format-links:
  - html
  - format: academic-typst
    text: PDF
    icon: file-pdf
abstract: |
  **Background**: Greater, more extensive collaboration is increasingly needed
  in the modern research environment, often to the level of team-based science.
  Combined with this need is the emphasize on doing more open science. However,
  there seems to be less attention on the overlap between open science and
  collaboration.

  **Objectives**: To conduct a scoping review on open collaborative practices used 
  across academia, including methods, tools, and resources.

  **Eligibility criteria**: We included any document that discussed, in any way,
  open collaboration practices, only those in English, and only those that were
  open access. We searched 5 years from whenever we ran a search.

  **Sources of evidence**: We used the OpenAlex database to search for scholarly
  works, which includes a large number of sources. Our search period was
  between 2019-02-28 to 2024-02-29.

  **Charting methods**: All three authors participated in the search, extraction,
  and charting process. As no standard charting form exists for our purposes,
  we developed our own template.

  **Results**: TBD

  **Conclusions**: TBD
execute: 
  echo: false
  message: false
  warning: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE)
library(stringr)
library(here)
source(here("R/openalex-search.R"))
search_terms_inline <- glue::glue("({search_terms()$title_search}[title]) AND {search_terms()$general_search}")
reviewed_records <- targets::tar_read(reviewed_article_records)

list_refs <- bib2df::bib2df(here("doc/reviewed-records.bib")) |>
  dplyr::filter(CATEGORY != "COMMENT") |>
  glue::glue_data("@{BIBTEXKEY}") |>
  str_c(collapse = ";") |>
  append("[", 0) |>
  append("]") |>
  str_c(collapse = "")
```

## Background

The time when a single scientist could work alone in a lab and make
meaningful progress in their field is long gone. Scientific research in
the modern era, especially in certain fields, requires working with
other people, oftentimes a large number of people [@Forscher2022].
Research questions are increasingly more challenging to answer as they
require more time, resources, expertise, and personnel in order to
meaningfully answer them.

With the current needs and limitations when doing scientific research,
collaborations range from those within a single group or lab, to a few
researchers across different departments in the same institution, to
many researchers and personnel spread across multiple research centers.
Collaborations are needed to address basic limitations of working with
small or constrained populations, to sharing valuable but expensive
equipment, or to assisting with technical expertise or domain knowledge.

At the same time, there is also a higher need for science to be more
open, higher quality, and more rigorous [@Cole2024; @Umbach2024;
@Korbmacher2023], which necessitates that researchers collaborate with
more researchers and with a more diverse group of researchers to fulfill
these needs. Open scientific practices, such as open data, open source,
and open materials, aim to improve the reliability and accessibility of
scientific output to increase the societal impact of the results.
There's greater attention and scrutiny towards verification of results,
a core tenet of science, through replication and reproducibility. In
many ways, the "replication crisis" in science has been a driving force
behind the push for more open science and "team science" practices that
necessitate more collaboration [@Korbmacher2023].

There are many claimed benefits to openness in science, including in
collaboration [@Forscher2022; @Kavanagh2021]. For instance:

-   Greater transparency and accountability, because the built-in
    openness makes it easy to see what was done, how, and why.
-   Better reproducibility and inspectability, because everything can be
    easily viewed and can be verified, which is a key tenet of science.

The intersection between greater collaboration and more openness
highlights key challenges researchers face. How can we effectively
collaborate together in a way that fulfills the demand for more
openness, transparency, and accountability, while at the same time also
not substantially impacting our productivity (or ideally improving it).
How can we collaborate in an open and transparent way? What are the best
practices and tools we can use? What is an ideal collaborative workflow
and how close or how far are we from this ideal in reality?

From our own personal experience working in the field of health
research, we have varied and diverse ways of collaborating and of being
open. Often the most commonly used approach to collaboration is emailing
Word documents around to get feedback and using meetings to discuss and
agree on things. Rarely do we systematically examine whether we are
following best practices for collaboration nor do we collaborate in an
open way following the principles of open science. As for open science,
while it was first introduced in 1985 [@Chubin1985] and discussed in
academic outputs regularly since then, the actual practices of being
open in science have been minimal or largely confined to specific
sub-branches of openness like open access.

This lack of visible role-models or standards of practice for
effectively collaborating and actively working openly presents a clear
barrier to uptake of these practices. This has real consequences to
continued productivity and quality of scientific research, which is
increasingly apparent in the modern era. These consequences are even
more apparent when comparing to more technically-inclined communities
such as those in open source software. These communities have massively
benefitted from integrating more openness into their workflows, to the
extent that the vast majority of the world's digital infrastructure is
built on and runs open source software. This highlights a greater need
for a paradigm shift in how we researchers collaborate to be able to
effectively and efficiently tackle the challenges of the modern era.

Open collaboration has many potential benefits, of which most overlap
with the benefits of open science. For instance, open collaboration can
lead to:

-   Easier onboarding of new collaborators, since everything is open and
    easily findable.
-   Higher collaboration opportunities, because more openness means more
    visibility as more parts of the scientific workflow are
    disseminated.
-   Faster time to dissemination, because the built-in openness makes it
    easier to disseminate results.

Unfortunately, there are, to our knowledge, very few resources and
examples of how to integrate open science into collaborations and how to
make collaboration be open [@Korbmacher2023].

### Objectives

Our aim is to conduct a scoping review to provide an overview of current
practices of open collaboration in the academic world. For this scoping
review, we define open collaboration using the definition as found in
[@Forte2013]:

@@

@@

1.  Provide an overview of current practices of or opinions about
    research collaboration that follow basic open principles (e.g.,
    transparency, accessibility)

2.  Summarize existing online tools and resources available to improve
    open collaboration in research

@@

This is a scoping review, so we followed the framework described in
[@Arksey2005; @Levac2010] as well as the guidelines outlined in the
PRISMA-ScR statement [@Tricco2018].

@@

While we didn't change our search terms for the initial search and
extraction, we found that our search terms were not precise enough and
got many irrelevant scholarly works. The full R code to filter down the
search results is kept in the
[`R/exclusions.R`](https://github.com/science-collective/scoping-review/blob/main/R/exclusions.R)
file on the GitHub repository. For example, we extracted a large number
of scholarly works describing surgery such as "open wound", electronics
such as "open circuit", or environmental such as "open water" that we
had to post-process and exclude.

@@

We found `r nrow(reviewed_records)` source documents discussing open
collaboration in some way. These were `r list_refs`. Two of the
documents, while passed our abstract review, did not contain anything
relevant when we reviewed the full text [@Network2023a; @Yordanov2022],
so we did not continue with charting them.

@@

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level
[@Turoman2022], center level [@Bush2022; @Network2023], multi-center
level [@ManyPrimates2021; @Network2023] and field level: psychology
[@Alessandroni2022], spinal cord injury [@TorresEspin2021] and
neuroimaging [@Niso2022]. All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

@@

For center-level only one paper was identified [@Bush2022], providing
the example of a center of neuroimaging describing first how they
identified the needs for open collaboration, then the individual-level
practices that researchers should adapt and then the center-level
changes to promote reaching FAIR goals.

@@

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates [@ManyPrimates2021;
@Primates2019] and one from the Reproducibility Network [@Network2023].
In this organization level the most urgent needs are: 1) identifying the
needs for open collaboration and 2) creating an infrastructure that
allows those needs to be satisfied. Indeed, the Reproducibility Network
[@Network2023] describes a systematic review as an exploration of the
needs for 11 UKRN centers to implement open collaborative practives. In
this paper they specially focus on tools and resources among the 11
institutions and make clear the need for starting to utilize open
resources as a first step to adapt open collaborative practices. The
first paper from ManyPrimates [@Primates2019] describes the start of
another multi-center collaboration, albeit from different institutions.
It is interesting to observe that in this situation, the complex issue
is to start the collaboration itself, which was ignited thanks to a
symposium and an email chain among future collaborators. In contrast,
ManyPrimates could quickly develop an open collaborative structure with
relative ease in comparison to changing an already solid infrastructure,
like in the case of @Network2023.

>>>>

---
title: "Towards open collaboration in biomedical and health research: An open collaborative scoping review"
date: last-modified
author:
  - name: Mario Garcia
    orcid: "0000-0002-3376-9460"
    affiliations:
      - ref: ku
  - name: Daniel B. Ibsen
    orcid: "0000-0002-7038-4770"
    affiliations:
      - ref: sdca
      - ref: au
  - name: Luke W. Johnston
    orcid: "0000-0003-4169-2616"
    affiliations:
      - ref: sdca
      - ref: au
affiliations:
  - id: sdca
    name: Steno Diabetes Center Aarhus
    country: Denmark
    url: "https://www.stenoaarhus.dk/"
  - id: au
    name: Aarhus University
    country: Denmark
    url: "https://www.au.dk/"
  - id: ku
    name: Copenhagen University
    country: Denmark
    url: "https://www.ku.dk/"
format:
  html: default
  typst:
    columns: 2
    papersize: a4
    margin:
      x: 2cm
      y: 2cm
format-links:
  - html
  - format: typst
    text: PDF
    icon: file-pdf
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE)
library(stringr)
library(here)
source(here("R/openalex-search.R"))
search_terms_inline <- glue::glue("({search_terms()$title_search}[title]) AND {search_terms()$general_search}")
```

## Introduction

-   Open science and its role in research

-   Collaboration and its role in modern research

    -   collaboration spanning from lab, to research center to multiple
        centers
    -   collaboration needed to tackle limitations: small populations,
        lack of replication, etc
    -   re-inventing discovery: a new way of working in science

Scientific research now almost always requires working with other
people.

With the growing complexity and specialization in scientific practices
and methods, together with globalisation of health and environmental
issues, there is a great need for a paradigm shift in research
collaboration to be able to tackle these challenges.

-   Open collaboration - a combination of two themes

    -   definition
    -   big potential in science
    -   but there very few resources and examples of how to integrate
        open science into collaborations

We define open collaboration using the definition as found in
[@Forte2013]:

@@

With the increasing emphasis on and demand for science to be more open,
how we collaborate together is a key component to making science more
open from the start of any project. But how do we collaborate in an open
and transparent way? What are the best practices and tools we can use?
What is an ideal collaborative workflow and how close or far are we from
this ideal in reality?

### Aim

This scoping review will focus on current practices of open
collaboration and open science in relation to collaboration in the field
of biomedical and health research.

@@

-   Provide an overview of current practices of or opinions about
    research collaboration that follow basic open principles (e.g.,
    transparency, accessibility)

-   Summarize existing online tools and resources available to improve
    open collaboration in research

@@

This is a scoping review, so we followed the framework described in
[@Arksey2005, @Levac2010] as well as the guidelines outlined in the
PRISMA-ScR statement [@Tricco2018].

@@

While we didn't change our search terms for the initial search and
extraction, we found that our search terms were not precise enough and
got many irrelevant scholarly works. The full R code to filter down the
search results is kept in the
[`R/exlusions.R`](https://github.com/science-collective/scoping-review/blob/main/R/exclusions.R)
file on the GitHub repository. For example, we extracted a large number
of scholarly works describing surgery such as "open wound", electronics
such as "open circuit", or environmental such as "open water" that we
had to post-process and exclude.

@@

### Scoping review identifies only 11 papers discussing open collaboration

@@

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level (Turoman et
al) center level (Bush et al and Grange et al), multi-center level
(ManyPrimates et al and Grange et al) and field level: psychology
(Alessandorni et al), Spinal Cord injury (Torres-Espin et al) and
neuroimaging (Niso et al). All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

@@

For center-level only one paper was identified (Bush et al), providing
the example of a center of neuroimaging describing first how they
identified the needs for open collaboration, then the individual-level
practices that researchers should adapt and then the center-level
changes to promote reaching FAIR goals.

@@

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates et al (ManyPrimates
2019 and ManyPrimates 2021) and Grange et al. In this organization level
the most urgent needs are: 1) identifying the needs for open
collaboration and 2) creating an infrastructure that allows those needs
to be satisfied. Indeed, Grange et al describes a systematic review as
an exploration of the needs for 11 UKRN centers to implement open
collaborative practives. In this paper they specially focus on tools and
resources among the 11 institutions and make clear the need for starting
to utilize open resources as a first step to adapt open collaborative
practices. The first paper from ManyPrimates et al \[REF\] describes the
start of another multi-center collaboration, albeit from different
institutions. It is interesting to observe that in this situation, the
complex issue is to start the collaboration itself, which was ignited
thanks to a symposium and an email chain among future collaborators. In
contrast, ManyPrimates could quickly develop an open collaborative
structure with relative ease in comparison to changing an already solid
infrastructure, like in the case of Grange et al.
